---
title: "Most Harmful Storm Events Across the States for People and Economy"
author: "Sertinell"
date: "12 November 2015"
output: html_document
---
# Sipnosys
In this study we explore the impact of storm events in the life of the United
States of America inhabitants. We explore both, the impact in health and in
economy. To identify the most harmful events for health we compare the total
number of fatalities and injuries caused by each kind of event. To identify the
most harmful eventsin terms of economic impact we compare tha amount of 
property damage and crop damage. We found that only four type events account
for almost the 80% of the fatalities and more than 85% of the injuries.
Simlarly, the five most harmful type of events account for more than 97% of the
property damage and more than 84% of the crop damage.

# Data Processing

```{r}
require(data.table)
require(dplyr)
require(ggplot2)
require(scales)
```

### Load data
Download the file, read it using fread through a bzcat to uncompress it
into a *data.frame*.

```{r cache=TRUE}
filename <- "data/storms.csv.bz2"
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = filename)
stormdata <-fread(sprintf("bzcat %s | tr -d '\\000'", filename))
```

### Clean up data
Since all the questions refer to data across the states, I have dropped out all the columns related to location.

```{r}
notwanted <- c(1,4:7,9:11,14:20, 30:37)
stormdata <- select(stormdata, -notwanted)
```

Change the columns names to lower case and remove "\_"

```{r cache=TRUE}
names(stormdata) <- tolower(names(stormdata))
names(stormdata) <- sub("_", "", names(stormdata))
names(stormdata)
```

### Calculate the actual damage

It seems that the *propdmgexp* and the *cropdmgexp* are the exponents of the *propdmg* and *cropdmg* respectively. We are using those colums to generate an extra column with the absolute value.

```{r}
table(stormdata$cropdmgexp)
table(stormdata$propdmgexp)
```

Numbers are taken as exponent of scientific notation (10*x) and letters are taken as follows:

* B: Billions 10^12
* k|K : Kilo 10^3
* M|m : Millions 10^6
* H|h : Hecto 10^2

Other values and blanks are assigned 0.

```{r cache=TRUE}
values <- data.frame(
  letters = c("B", "k", "K", "m", "M", "H", "h"), 
  value =   c(  12,   3,   3,  6,   6 ,   2,  2) 
  );
for(letter in values$letters){
  stormdata$cropdmgexp[stormdata$cropdmgexp == letter] <-
    values$value[values$letters == letter]
  stormdata$propdmgexp[stormdata$propdmgexp == letter] <-
    values$value[values$letters == letter]
}

stormdata$cropdmgexp <- as.integer(stormdata$cropdmgexp)
stormdata$propdmgexp <- as.integer(stormdata$propdmgexp)
stormdata$cropdmgexp[is.na(stormdata$cropdmgexp)] <- 0
stormdata$propdmgexp[is.na(stormdata$propdmgexp)] <- 0

stormdata <- mutate(stormdata, propertydamage = propdmg * 10^propdmgexp, cropdamage = cropdmg * 10^cropdmgexp )

```

Drop old columns

```{r}
stormdata <- select(stormdata, -c(cropdmg, propdmg, cropdmgexp, propdmgexp))
```


### Clean up the *evtype* column
Extract the evtype column and inspect it to decide how to clean up this messy column. Get all the unique values and short it alphabetycally.

```{r}
evtype <- unique(stormdata$evtype)
evtype <- evtype[order(evtype)]
evtype <- tolower(evtype)
evtype
```

I have decided to keep only the most common *words* and consider every evtype that contains a certain word as an event of this word type.

Split all the evtype values by words to find the most common words longer than 2 characters:

```{r}
words <- unlist(strsplit(evtype,'[^a-zA-Z]', perl = TRUE))
words <- unique(words)
words <- words[order(words)]
words <- words[nchar(words) > 2]
```

Count the times every type appears:

```{r cache=TRUE}
stormdata$evtype <- tolower(stormdata$evtype)
wordtable <- lapply(words, FUN=function(x){
              sum(grepl(x, stormdata$evtype))
             })
wordtable <- data.frame(word = words, count = unlist(wordtable))
wordtable <- wordtable[order(-wordtable$count),]
head(wordtable, 20)
```

We can see that words like *heavy*, *high* and others appear on the top of our list. Drop them because they are too vague.

```{r}
dropwords <- c("heavy", "high", "strong", "extreme", "record", "light")
wordtable <- wordtable[-which(wordtable$word %in% dropwords),]
```

Now we need to run a loop and assign the *word* value to a new column called *cleanevent* to every event that matches each of the *words*. Every event is assigned a *cleanevent* value only once.

```{r cache=TRUE}
stormdata$cleanevent <- as.character("")
stormdata$cleanevent <- NA
for(event in wordtable$word)
  {
    stormdata$cleanevent[is.na(stormdata$cleanevent) &
                          grepl(event, stormdata$evtype) ] <- event
  }
```

We got *only* 85 levels and there is only 9 NA event. The NA values don't seem to be important.

```{r}
length(table(stormdata$cleanevent))
sum(is.na(stormdata$cleanevent))
stormdata[is.na(stormdata$cleanevent),]
```

## Which types of events are most harmful with respect to population health?

To show the impact of events in population health we need to consider the columns *injuries* and *fatalities*. Short the data according to those columns. It seems the worst recorded events were tornadoes and heatwaves.

```{r}
stormdata <- arrange(stormdata, desc(fatalities),
                     desc(injuries))
head(stormdata)
```

Now we can summarize the data to check the totals:

```{r}
databyev <- group_by(stormdata, cleanevent)
databyev <- summarise_each(databyev, funs(sum), fatalities, injuries)
databyev <- arrange(databyev, desc(fatalities), desc(injuries))
head(databyev)
```

Check the percentage of total victims per event:

```{r}
databyev <- mutate(databyev, 
                   fatalitiespercent  = fatalities/sum(fatalities), 
                   injuriespercent = injuries/sum(injuries) )
databyev <- mutate(databyev, 
                   cumsumfat  = cumsum(fatalitiespercent), 
                   cumsuminj = cumsum(injuriespercent) )
head(databyev, 10)
```

## Which types of events have the greatest economic consequences? 

To show the impact on economy we inspect the columns *propertydanage* and *cropdamage*.

```{r}
stormdata <- arrange(stormdata, desc(propertydamage),
                     desc(cropdamage))
head(stormdata)
```

It seems like floods, storms and hurricanes are the most harmful events in terms of economical impact. Summarize the data to get the totals.

```{r}
databyeveco <- group_by(stormdata, cleanevent)
databyeveco <- summarise_each(databyeveco, funs(sum), propertydamage, cropdamage)
databyeveco <- arrange(databyeveco, desc(propertydamage), desc(cropdamage))
head(databyeveco)
```

Check the percentage of damage for every event:

```{r}
databyeveco <- mutate(databyeveco, 
                   propertydamagepercent  = propertydamage/sum(propertydamage),
                   cropdamagepercent = cropdamage/sum(cropdamage) )
databyeveco <- mutate(databyeveco, 
                   cumsumprop  = cumsum(propertydamagepercent), 
                   cumsumcrop = cumsum(cropdamagepercent) )
head(databyeveco)
```

#Results 

### Effects on Population Health

Four events account for almost the 80% of the fatalities and more than 85% of the injuries. Those events are:

1. Tornadoes
2. Heatwaves
3. Wind
4. Flood
5. Lightning
```{r}
head(databyev, 10)
```


Plot a barplot:

```{r}
temp <- head(databyev, 5)
temp <- select(temp, cleanevent, fatalitiespercent, injuriespercent)
temp$cleanevent <- as.factor(temp$cleanevent)
levels(temp$cleanevent)<- c("flood", "heatwave", "lightning",
                            "tornado", "wind")
temp$cleanevent <- reorder(temp$cleanevent, desc(temp$fatalitiespercent), min)
temp <- melt(temp, id.vars = "cleanevent", variable.name = "variable")
levels(temp$variable) <- c("Fatalities", "Injuries")
ggplot(temp, aes(x=cleanevent, y=value))+
  geom_bar(stat = "identity") +
  facet_grid( . ~ variable) +
  ggtitle("Most Harmful Events for Population Health")+
  xlab("Event Type") +
  ylab("Percentage of total")+
  scale_y_continuous(labels = percent) 
```

### Effects on Economy

The five most harmful events account for more than 97% of the property damage and more than 84% of the crop damage. Those events are:

1. Floods
2. Hurricanes
3. Storms
4. Wind
5. Tornadoes

```{r}
head(databyeveco, 10)
```

Plot a barplot:

```{r}
temp <- head(databyeveco, 5)
temp <- select(temp, cleanevent, propertydamagepercent, cropdamagepercent)
temp$cleanevent <- as.factor(temp$cleanevent)
levels(temp$cleanevent)[5] <- "wind" 
temp$cleanevent <- reorder(temp$cleanevent, desc(temp$propertydamagepercent), min)
temp <- melt(temp, id.vars = "cleanevent", variable.name = "variable")
levels(temp$variable) <- c("Property Damage", "Crop Damage")
ggplot(temp, aes(x=cleanevent, y=value))+
  geom_bar(stat = "identity") +
  facet_grid( . ~ variable) +
  ggtitle("Most Harmful Events for Economy")+
  xlab("Event Type") +
  ylab("Percentage of total")+
  scale_y_continuous(labels = percent) 
```
